MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-06-18 latex/paper/sections/05-7series_architecture.tex
packing edits
start placement results
results discussion

In the 7-Series architecture, the term \emph{CLB} (Configurable Logic Block) refers to a \emph{CLB Tile} that contains two \emph{SLICE} Sites. 
Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.
\begin{itemize}
\item Each \texttt{SLICEL} has a set of BELs including eight \texttt{LUT}s, eight \texttt{FF}s, and one \texttt{CARRY4} adder. The \texttt{LUT} BELs in a \texttt{SLICEL} can only host \texttt{LUT} Cells. 
\item The \texttt{SLICEM} includes all the features of a \texttt{SLICEL} but its \texttt{LUT} BELs can host both \texttt{LUT} Cells, which are asynchronous ROM elements, or \texttt{RAM32M} Cells, which are synchronous 32-deep RAM elements. These cells are also referred to as \emph{Distributed RAM} in the Xilinx documentation.
    These cells can offer an alternative to the larger, more dedicated 18K-36K \texttt{RAMB18E1} cells when RAM resources are highly utilized or when the design inheritly demands homogeneously distributed local memory. 
\end{itemize}
In a typical 7-Series device, approximately 75\% of the SLICE Sites are \texttt{SLICEL}s and 25\% are \texttt{SLICEM}s. 
A single CLB Tile can therefore host either two \texttt{SLICEL}s or one \texttt{SLICEL} and one \texttt{SLICEM}.
To simplify the problem space, however, we will only consider \texttt{SLICEL}s for general logic and use the dedicated \texttt{RAMB18E1} cells to implement RAM elements.

@@

\textbf{LUTs} \quad
Combinational logic is universal to all HDL designs. 
As the their name suggests, a Look‑Up Table (LUT) map an input value to an output value. 
LUTs facilitate combinational logic by acting as tiny asynchronously‑accessed ROMs whose contents are fixed when the FPGA is programmed.  
For any boolean function, the synthesizer precalculates the boolean output to every possible input combination and stores the resulting truth table into a LUT's static memory. 
The inputs are then essentially treated as an address space that maps to a data value space in an asynchronous ROM. 
No explicit logic gates like NAND or XNOR are synthesized, contrary to what newcomers might expect from a "Field Programmable Gate Array".

>>>>

In the 7-Series architecture, the term \emph{CLB} (Configurable Logic Block) refers to a \emph{CLB Tile} that contains two \emph{SLICE} Sites. 
Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.
\begin{itemize}
\item Each \texttt{SLICEL} has a set of BELs including eight \texttt{LUT}s, eight \texttt{FF}s, and one \texttt{CARRY4} adder. The \texttt{LUT} BELs in a \texttt{SLICEL} can only host \texttt{LUT} Cells. 
\item The \texttt{SLICEM} includes all the features of a \texttt{SLICEL} but its \texttt{LUT} BELs can host both \texttt{LUT} Cells, which are asynchronous ROM elements, or \texttt{RAM32M} Cells, which are synchronous 32-deep RAM elements. These cells are also referred to as \emph{Distributed RAM} in the Xilinx documentation.
    These cells can offer an alternative to the larger, more dedicated 18K-36K \texttt{RAMB18E1} cells when RAM resources are highly utilized or when the design inheritly demands homogeneously distributed local memory. 
\end{itemize}
In a typical 7-Series device, approximately 75\% of the SLICE Sites are \texttt{SLICEL}s and 25\% are \texttt{SLICEM}s. 
A single CLB Tile can therefore host either two \texttt{SLICEL}s or one \texttt{SLICEL} and one \texttt{SLICEM}.
However, to simplify the problem space, we will only consider \texttt{SLICEL}s for general logic and use the dedicated \texttt{RAMB18E1} cells to implement RAM elements.

@@

\textbf{LUTs} \quad
Combinational logic is universal to all HDL designs. 
As the their name suggests, a Look‑Up Table (LUT) map an input value to an output value. 
LUTs facilitate combinational logic by acting as tiny asynchronously‑accessed ROMs whose contents are fixed when the FPGA is programmed.  
For any boolean function, the synthesizer precalculates the boolean output to every possible input combination and stores the resulting truth table into a LUT's static memory. 
The inputs are then essentially treated as an address space that maps to a data value space in an asynchronous ROM. 
No explicit logic gates like NAND or XNOR are synthesized, contrary to what the name \emph{Field Programmable Gate Array} might suggest.

### 2025-06-18 latex/paper/sections/10-placement.tex
results discussion

Shown in Listing \ref{lst:sa_outer} is a simplified pseudocode for the outer-most loop of our placer's SA algorithm. 
First, we randomly place all \texttt{SiteInst}s across the \texttt{device}.
Then, we initialize a simple geometric cooling schedule: 
\begin{equation}
    \label{coolingSchedule}
    T_{n+1} = \alpha T_n
\end{equation}
with initial temperature \(T_0=10000\), cooling rate \(\alpha=0.98\), and \(n = \{0, 1, 2, ..., 300\}\) passes.

@@

In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection. 
Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
Movement approaches that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" moves.

>>>>

Shown in Listing \ref{lst:sa_outer} is a simplified pseudocode for the outer-most loop of our placer's SA algorithm. 
First, we randomly place all \texttt{SiteInst}s across the \texttt{device}.
Then, we initialize a simple geometric cooling schedule: 
\begin{equation}
    \label{coolingSchedule}
    T_{n+1} = \alpha T_n
\end{equation}
with initial temperature \(T_0=10000\), cooling rate \(\alpha=0.98\), and \(n = \{0, 1, 2, ..., 300\}\) passes.
Hill-climbing moves can be accepted via the probability function \(P()\): 
\begin{equation}
    \label{moveAcceptance}
    P(s, s^*, T_n) = exp(\frac{E(s) - E(s^*)}{T_n})
\end{equation}
where \(E(s^*)\) is the energy of the proposed state, \(E(s)\) is the energy of the current state, and \(T_n\) is the current global temperature at \(n\) passes. 
If \(P(s, s^*, T_n) < random(0, 1)\), then \(s \leftarrow s^*\).
Recall that this probability is only evaluated for hill-climbing moves, thus cost difference \(E(s) - E(s^*)\) is always negative.
In summary, the higher the cost difference and temperature, the higher the probabilty of hill-climb acceptance.
On the other hand, all hill-descending moves are simply accepted outright.

@@

In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection. 
Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" moves and are akin to physics simulators modelling interactions between atoms and molecules.

### 2025-06-18 latex/paper/sections/11-placement_results.tex
results discussion

Here, we present test results for a basic SA placer as well as test results for four simple variations as listed below. 
Note that midpoint simply means centroid.

\begin{itemize}
    \item \texttt{PlacerGreedyRandom}: all undirected random moves with greedy acceptance
    \item \texttt{PlacerGreedyMidpoint}: all directed centroid moves with greedy acceptance
    \item \texttt{PlacerAnnealRandom}: all undirected random moves with annealing acceptance \textbf{(the basic SA algorithm)}
    \item \texttt{PlacerAnnealMidpoint}: all directed centroid moves with annealing acceptance
    \item \texttt{PlacerAnnealHybrid}: 50\% random - 50\% centroid moves with annealing acceptance
\end{itemize}

@@

Shown in figure \ref{fig:placers_overlay} are the HPWL curves over number of passes for each of the five placer variants. 
The Midpoint-exclusive placers \texttt{PlacerGreedyMidpoint} and \texttt{PlacerAnnealMidpoint} appear

>>>>

@@

Now, we present test results for a basic SA placer as well as test results for four simple variations as listed below. 
Note that midpoint simply means centroid.

\begin{itemize}
    \item \texttt{PlacerGreedyRandom}: all undirected random moves with greedy acceptance (always zero temperature)
    \item \texttt{PlacerGreedyMidpoint}: all directed centroid moves with greedy acceptance (always zero temperature)
    \item \texttt{PlacerAnnealRandom}: all undirected random moves with annealing acceptance \textbf{(the basic SA algorithm)}
    \item \texttt{PlacerAnnealMidpoint}: all directed centroid moves with annealing acceptance
    \item \texttt{PlacerAnnealHybrid}: \textbf{Initial:} 50\% random - 50\% centroid moves with annealing acceptance. \textbf{At Near-Zero Temperature:} 100\% centroid moves with annealing acceptance.
\end{itemize}

((Talk about the initial cost.))

Shown in figure \ref{fig:placers_overlay} are the HPWL curves over number of passes for each of the five variants. 
\texttt{PlacerGreedyMidpoint} (in red) appears to perform the worst by a considerable margin with a final HPWL cost of 590578.
This is likely due to the lack of randomness in moves resulting in a lack of diversity in movement proposals combined with the greediness of the movement acceptance that prevents the system from escaping local minima.
This is supported by its HPWL curve (in red) showing a sharp drop in cost that quickly asymptotes just below the 600000 mark, suggesting that the system crystallized very quickly into a local minimum and was unable to climb out of it.

Next is \texttt{PlacerAnnealMidpoint} with a final cost of 466855, which like the previous placer, lacks the diversity of random moves and contributes to faster crystallization into local minima.
However, this placer likely performed better than the previous as there is still some randomness in the move acceptance evaluation, which gives way to occasional hill-climbing. 
Keep in mind that centroid moves do not necessarily decrease system cost, as the centroid move will decrease the HPWL cost of the nets on connected to the current \texttt{SiteInst}, but potentially increase the cost of another group of nets should a swap be evaluated, which can potentially result in a total increase in system cost.
